Wide-Field Effects in Redshifted 21~cm Power Spectra

### 2015-03-18 ms.tex
Modify abstract, adjust all sections accordingly
From ac1c778:
============

In ms.tex,

(*) Modify abstract and adjust all sections accordingly

Foreground emission is currently the primary limitation to detection of redshifted H~{\sc i} emission from the epoch of reionization. Modern radio telescopes that target this cosmological signal are typically wide-field instruments. Through modeling of delay spectra measured between antenna pairs, it has recently emerged that wide-field measurements imprint a characteristic {\it pitchfork}-shaped signature in this Fourier domain. This predicted feature is characterized by enhanced power from foreground emission mapped to regions near the horizon and plays a significant role in determining the contamination of the cosmological H~{\sc i} signal. With MWA data sensitivity improved by coherently averaging 12 independent snapshots aligned in local sidereal time across different observing nights, we confirm the prediction with a signal-noise ratio $>10$.

@@

The primary challenge to detection of cosmological redshifted H~{\sc i} from the EoR comes from continuum emission from Galactic and extragalactic foreground objects, relative to which the desired signal is $\sim 10^4$ times weaker \citep[see, e.g.,][]{dim02,zal04,fur06,ali08,ber09,ber10,gho12}. Since the foregrounds and the signal have inherent differences in spatial isotropy and spectral smoothness \citep{mor04,mor06,bow09,liu11,par12b,dil13,pob13}, a detailed characterization of foreground emission has become essential \citep{bow09,liu09,dat10,liu11,mor12,tro12,pob13,dil14,liu14a,liu14b,thy13,thy15}.

A recent study by \citet{thy15} using instrument and all-sky foreground models has shed new light on the significant role played by the interplay between foreground emission and the nature of wide-field measurements, typical of most modern EoR experiments. This paper presents a follow-up analysis of that study. By using a larger volume of MWA data obtained independently over different nights pointed at the same region of sky, we confirm with high significance the presence of certain key characteristics (referred to as the {\it pitchfork} signature) of wide-field measurements predicted in the preceding study.

\S\ref{sec:wide-field} gives an overview of the role of wide-field measurements in the delay spectral domain and the predicted {\it pitchfork} signature. In \S\ref{sec:MWA}, we describe the data analysis procedure used to improve the dynamic range of delay spectra of MWA data. We describe the results and confirm the presence of the predicted wide-field effects in \S\ref{sec:results}. \S\ref{sec:summary} summarizes our findings.

@@

In wide-field measurements, the steep rise in subtended solid angle near the horizon for a fixed delay bin size significantly enhances the integrated emission near the horizon delay limits. This is found to be true for diffuse emission even on wide antenna spacings because their foreshortening towards the horizon makes them sensitive to large angular scales that match the inverse of their foreshortened lengths.

Typically, an antenna is most sensitive towards its primary field of view relative to the rest of the sky, which in conjunction with the aforementioned wide-field effects results in a characteristic {\it pitchfork} signature in the delay spectrum.

Although there is marginal evidence for presence of this feature in the {\it zenith} snapshot presented in \citet{thy15}, the high level of thermal noise prevented a robust confirmation. In this paper, we confirm the {\it pitchfork} signature using deeper data from the MWA.

@@

In a recent study, through the use of models for an all-sky foreground and the instrument, the role of wide-field nature of modern radio interferometric measurements, hitherto unpredicted, has emerged. It is referred to as the {\it pitchfork} signature in the delay spectral domain, and is characterized by significant foreground power in the primary field of view and an enhancement in power near the horizon. The latter is due to the highly non-linear mapping between geometric delays and subtended solid angles besides an increase in sensitivity to larger scales caused by foreshortening of baseline lengths along these directions. This feature could not be confirmed in individual MWA snapshot observations due to the high level of thermal fluctuations.

By averaging 12 independent snapshots tightly aligned in LST from MWA observations, we improve the dynamic range in delay power spectra by a factor $\gtrsim$~10 while maintaining the coherency of foreground contributions in the observed {\it visibilities}. Foreground power in delay bins near the horizon is distinctly visible with a signal-noise ratio $>10$. In addition, there is a close resemblance in the morphologies of delay power spectra from data and models. Together, they allow us to make a robust confirmation of the nature of wide-field measurements in EoR experiments.

>>>>

While probing H~{\sc i} in the epoch of reionization (EoR), through modeling of delay spectra of measurements between antenna pairs, it has recently emerged that the foreground structure, the nature of wide-field measurements and the antenna aperture imprint a characteristic {\it pitchfork}-shaped signature. By improving MWA data sensitivity through coherent averaging of 12 independent snapshots aligned in local sidereal time across different observing nights, we provide the first confirmation of the prediction with a signal-noise ratio $>10$. This feature is characterized by enhanced power from foreground emission in Fourier modes mapped to regions near the horizon, and is significant for EoR studies because Fourier modes considered to be most sensitive for the cosmological H~{\sc i} signal are very susceptible to contamination from these regions.

% Foreground emission is currently the primary limitation to detection of redshifted H~{\sc i} emission from the epoch of reionization. Modern radio telescopes that target this cosmological signal are typically wide-field instruments. Through modeling of delay spectra measured between antenna pairs, it has recently emerged that wide-field measurements imprint a characteristic {\it pitchfork}-shaped signature in this Fourier domain. This predicted feature is characterized by enhanced power from foreground emission mapped to regions near the horizon and plays a significant role in determining the contamination of the cosmological H~{\sc i} signal. With MWA data sensitivity improved by coherently averaging 12 independent snapshots aligned in local sidereal time across different observing nights, we confirm the prediction with a signal-noise ratio $>10$.

% Foreground emission is currently the primary limitation to detection of redshifted H~{\sc i} emission from the epoch of reionization. Modern radio telescopes that target this cosmological signal are typically wide-field instruments. Through modeling of delay spectra measured between antenna pairs, it has recently emerged that wide-field measurements imprint a characteristic {\it pitchfork}-shaped signature in this Fourier domain. This predicted feature is characterized by enhanced power from foreground emission mapped to regions near the horizon and plays a significant role in determining the contamination of the cosmological H~{\sc i} signal. With MWA data sensitivity improved by coherently averaging 12 independent snapshots aligned in local sidereal time across different observing nights, we confirm the prediction with a signal-noise ratio $>10$.

@@

The primary challenge to detection of cosmological H~{\sc i} from the EoR comes from continuum emission from Galactic and extragalactic foreground objects, relative to which the desired signal is $\sim 10^4$ times weaker \citep[see, e.g.,][]{dim02,zal04,fur06,ali08,ber09,ber10,gho12}. Since the foregrounds and the signal have inherent differences in spatial isotropy and spectral smoothness \citep{mor04,mor06,bow09,liu11,par12b,dil13,pob13}, a detailed characterization of foreground emission has become essential \citep{bow09,liu09,dat10,liu11,mor12,tro12,pob13,dil14,liu14a,liu14b,thy13,thy15}.

A recent study by \citet{thy15} using instrument and all-sky foreground models has shed new light on the significant role played by the interplay between foreground emission and the nature of wide-field measurements, typical of most modern EoR experiments and was noted to cause significant contamination in Fourier modes considered to be most sensitive for the cosmological H~{\sc i} signal. This paper presents a follow-up analysis of that study. By using a larger volume of MWA data obtained independently over different nights pointed at the same region of sky, we confirm with high significance the presence of certain key characteristics (referred to as the {\it pitchfork} signature) of wide-field measurements predicted in the preceding study.

\S\ref{sec:wide-field} is an overview of the role of wide-field measurements in the delay spectral domain and the predicted {\it pitchfork} signature. \S\ref{sec:MWA} describes the analysis of MWA data used to improve the dynamic range of the delay spectra. \S\ref{sec:results} describes the results and confirms the presence of the predicted wide-field effects. \S\ref{sec:summary} summarizes our findings.

@@

In wide-field measurements, the steep rise in subtended solid angle near the horizon for a fixed delay bin size significantly enhances the integrated emission near the horizon delay limits. This is found to be true for diffuse emission even on wide antenna spacings because their foreshortening towards the horizon makes them sensitive to large angular scales that match the inverse of their foreshortened lengths. Typically, an antenna is most sensitive towards its primary field of view relative to the rest of the sky, which in conjunction with the aforementioned wide-field effects results in a characteristic {\it pitchfork} signature in the delay spectrum.

Detailed modeling presented in \citet{thy15} has shown this feature to be a leading source of contamination of the Fourier modes most sensitive to the EoR H~{\sc i} signal. Although there is marginal evidence for presence of this feature in the {\it zenith} snapshot presented in their study, the high level of thermal noise prevented a robust confirmation. In this paper, we confirm the {\it pitchfork} signature using deeper data from the MWA.

@@

In a recent study, through the use of models for an all-sky foreground and the instrument, the role of wide-field nature of modern radio interferometric measurements, hitherto unpredicted, has emerged. It is referred to as the {\it pitchfork} signature in the delay spectral domain, and is characterized by significant foreground power in the primary field of view and an enhancement in power near the horizon. The latter is due to the highly non-linear mapping between geometric delays and subtended solid angles besides an increase in sensitivity to larger scales caused by foreshortening of baseline lengths along these directions. Owing to its localization in delay spectral domain, it is a leading cause of contamination in Fourier modes most sensitive to the cosmological H~{\sc i} signal. This feature could not be confirmed in individual MWA snapshot due to the high level of thermal fluctuations.

By averaging 12 independent LST-aligned snapshots from MWA observations, we improve the dynamic range in delay power spectra by a factor $\gtrsim$~10 while maintaining the foreground contributions in the observed {\it visibilities} coherent. Foreground power in delay bins near the horizon is distinctly visible with a signal-noise ratio $>10$. Further, there is a close resemblance in the morphologies of delay power spectra from data and models. Together, they allow us to make a robust confirmation of the nature of wide-field measurements in EoR experiments.
